For Finitary Induction-Induction, \\ Induction is Enough

### 2020-05-03 paper/universaliit.tex
stuff

We construct the theory of IIT signatures in the following steps:
\begin{enumerate}
\item We view the theory of signatures as a type theory, and we define
  its untyped syntax as mutual inductive types together with typing
  judgments given by inductive relations on the untyped syntax. Then
  the syntax $\I : \SignAlg$ is constructed using those untyped terms
  for which the typing relation holds.
\item We construct $\ll\blank\rr:\SignMor\,\I\,M$ for arbitrary
  $M : \SignAlg$, by:
  \begin{enumerate}
  \item defining a relation $\blank \sim \blank$ between the well-typed syntax
    and a given model. The idea is that given a syntactic context $\GG$ and a
    semantic context $\model{\GG}$ of the model $M$, we have
    $\GG \sim \model{\GG}$ if and only if $\ll\GG\rr
    = \model{\GG}$, and similarly for types, terms, and substitutions;
    % semantic context $\model{\GG}$ of
    % the model $M$ if and only if $\GG$ relates to $\rec\,{\GG}$;
    % enjoyed by the initial morphism from the
    % syntax to a given model;
  \item showing that this relation is functional.
    \end{enumerate}
  \item Proving the uniqueness of this morphism by showing that any morphism
    $f:\SignMor\,\I\,M$ satisfies the relation. For example, for any syntactic context
    $\GG$ we have $\GG\sim f\,\GG$.
      % TODO donner composantes sur les contextes
\end{enumerate}
The next sections detail each of these steps.

@@

Crucially, we do not have conversion relations for typed syntax, nor
do we need to use quotients when giving the syntactic model. This is
possible because there are no $\beta$-rules in the theory of
signatures. Hence, we consider only normal terms in the untyped
syntax, and define weakening and substitution by recursion.  Avoiding
quotients is important for two reasons. First, it greatly simplifies
formalisation. Second, we aim to reduce the theory of signatures only
to inductive types, thus making \Cref{TODO} stronger.

@@

\subsubsection{Untyped syntax and its properties}

@@

\subsubsection{Typing judgments and their properties}

\begin{definition}[Typing judgements]\label{def:typing_judgments}
  The typing judgments are defined as the following inductive type indexed over the
  untyped syntax:
\begin{alignat*}{5}
  & \rlap{(1) Substitution calculus} \\[0.5em]
  & \blank \vdash && : \Conu\ra \Set \\
  & \blank \vdash \blank  && : \Conu\ra \Tyu\ra\Set \\
  & \blank \vdash \blank \invar \blank  && : \Conu\ra \N \ra \Tyu\ra \Set \\
  & \blank \vdash \blank \in \blank  && : \Conu\ra \Tmu\ra \Tyu\ra \Set \\
  & \blank \vdash \blank \Rightarrow \blank  && : \Conu\ra \Subu \ra \Conu\ra \Set \\
  % & \Subpre  && : \Set \\
  & \emptyw && : \emptyu \vdash \\
  & \epsilonw && : \GG\vdash \epsilonu \Rightarrow \emptyu \\
  & \blank\extw\blank && : ( \GG\vdash )  \ra  (\GG\vdash A)  \ra
  \GG\extu A \vdash \\
  & \consw && :
    (\GD \vdash) \ra
    (\GG\vdash \sigma \Rightarrow\GD)\ra
    ( \GD \vdash A) \ra
    ( \GG \vdash t \in A[\sigma]) \ra
    \GG \vdash \sigma\consu t \Rightarrow \GD\extu A
   \\
  & \varw  && : (\GG \vdash n \invar A) \ra \GG \vdash \varu n \in A \\
  & \vzw  && : (\GG\vdash)\ra(\GG\vdash A)\ra\GG \extu  A \vdash \mathsf{0} \invar \wku \, A \\
  & \vsw  && : (\GG\vdash)\ra(\GG\vdash A)\ra(\GG \vdash n \invar A) \ra (\GG \vdash B) \ra \GG \extu  B
  \vdash \mathsf{S}\,n \invar \wku \, A \\
  & \rlap{(2) Universe} \\[0.5em]
  & \Uw && : (\GG \vdash)\ra \GG \vdash \Uu \\
  & \Elw && : (\GG \vdash)\ra (\GG \vdash a \in \Uu)\ra\GG \vdash \Elu\,a \\
  & \rlap{(3) Inductive parameters} \\[0.5em]
\  & \Piw && :
    (\GG \vdash)\ra (\GG \vdash a \in \Uu)\ra(\GG \extu
  \Elu\,a \vdash B)\ra\GG \vdash \Piu \, a \, B \\
  & \appw && :
    (\GG \vdash)\ra (\GG \vdash a \in \Uu)\ra(\GG \extu
    \Elu\,a \vdash B)
    \\
    & &&
    \ra(\GG \vdash t \in \Piu \, a \, B )
    \ra
    (\GG \vdash u \in \Elu\,a)
    \ra
    \GG \vdash t \appu u \in  B [0 := u] \\
  & \rlap{(4) External parameters} \\[0.5em]
  & \Pimw && :
    (T:\Set)\ra(A : T\ra \Tyu)\ra(\GG \vdash)\ra
    ((t : T)\ra\GG\vdash A\,t) \ra
    \GG \vdash \Pimu \,T\,A
    \\
    & \appmw  && :
    (T:\Set)\ra(A : T\ra \Tyu)\ra(\GG \vdash)\ra
    ((t : T)\ra\GG\vdash A\,t)
    \\
    & &&
    \ra(\GG \vdash t \in \Pimu \,T\,A)
    \ra (u : T) \ra \GG \vdash t\,\appmu\,u \in A\,u
  \end{alignat*}
\end{definition}
There is possibility of redundancy in the arguments of the
constructors. Here, we are ``{paranoid}'' (nomenclature from
\cite{paranoid}), so that we get more inductive hypotheses when
performing recursion.

@@

\subsubsection{The syntax as a signature algebra}

@@

\subsection{Relating the Syntax to a Model}
\label{sec:streicher}

@@

xxx

Given an $M : \SignAlg$, we define the functional relation satisfied by
the initial morphism $\rec : \mathsf{S}\to M$ by recursion on the typing judgments.
% As these judgments are propositions, we will sometimes
If $\GG$ is a context in $S$ and $\model{\GG}$ is a semantic context
(i.e.\ a context of the model $M$), we want to define a type
$\GG\sim \model{\GG}$ equivalent to $\rec\,\,\GG=\model{\GG}$. Of
course, at this stage, $\rec$ is not available yet since the point of
defining this relation is to construct $\rec$ in the end.

For a type $A$ in a context $\GG$, we want to define a relation
$A\sim \model{A}$ that is equivalent to $\rec\,\,A=\model{A}$.
% where $\model{A}$ is a
% semantic type in a semantic context $\model{\GG}$.
For this equality to make sense, the semantic type $\model{A}$ must live
in the semantic context $\rec\,\,\GG$. But again, $\rec$ is not
yet available at this stage. Exploiting the expected equivalence between
$\GG\sim\model\GG$ and $\rec\,\,\GG=\model\GG$, we may consider defining $A\sim\model{A}$ under the
hypotheses that $\model{A}$ lies in a semantics context $\model\GG$ which is
related to $\GG$. Then, the type of the relation for types is
\[
  (\GG : \syn\Con)\ra (A : \syn\Ty\,\syn\GG) \ra
  (\model\GG : \model\Con)
  \ra
  (\GG \sim \model\GG)
  \ra
  (\model{A}:\model\Ty\,\model\GG)
  \ra
  \Set
\]
Note that the relation on contexts
must be defined mutually with the relation on types (see for example the case of
context extension), but here, the relation on contexts appears as the type of an
argument of the relation on types.
% We leave this possibility aside,
We want to avoid using such recursive-recursive
definitions as they are not allowed by the elimination principles of indexed inductive types, so we instead just remove the hypothesis
$\GG\sim\model\GG$ from the list of arguments.  We proceed similarly for
terms and substitutions. Actually, this removal is not without harm. For
example, consider relating the empty substitution
$\GG\vdash \epsilonu \Rightarrow \emptyu $ to a semantic substitution
$\model\sigma : \model\Sub\,\model\GG\,\model\GD$. We would like to assert
that $\model\sigma$ equals the empty semantic substitution $\model\epsilon$, but
this is not possible because typechecking requires that $\model\GD$ is the
empty semantic context. This is precisely what was ensured by the hypothesis
$\syn\cdot \sim \model\GD$ we removed.  Our way out here is to state that $\model\sigma$
is related to the empty substitution if the target semantic context $\model\GD$
is empty, and, acknowledging this equality, if $\model\sigma$ is the empty
substitution.
% The idea is that we will first prove that there is atmost one semantic type
% which relates to the syntactic one, and then later we will provide such a
% semantic type under the hypothesis that the contexts are related.

Let us mention another possible solution for avoiding recursion-recursion:
defining $A\sim\model{A}$ so that it is equivalent to $(e:\rec\,\GG
= \model{\GG})\times (\rec\,A = _{e\#}\model{A})$.
% whereas the right hand side is of type $\model\Ty\model{\GG}$.
% For a substitution $\sigma$ between contexts $\GG$ and $\GD$,
% a first idea consists in defining a relation $\sigma \sim_{\blank} \model{\sigma} : \GG
% \sim \model\GG \to \Set $ where $\model{A}$ is a semantic type in the
% semantic context $\model{\GG}$. We define the relation on types under the hypothesis
% that the contexts are already related.
% it is natural to define the relation with a semantic type $\model{A}$ in a
% semantic context $\model{\GG}$ such that it is equivalent to  $(\rec\,\GG = \model{\GG})\times (\rec\,A = \model{A})$ (note that the first equality is
% required for the second one to be well-typed).
% The discussion is similar for terms.
% Then, the ultimate goal is to prove that
% $\sum_{\model{\GG}}\GG\sim\model{\GG}$ and
% $\sum_{\model{\GG}}({\model{A}:\model{\Ty}\,\model{\GG}})\times(A\sim\model{A})$
% are contractible, i.e., have a unique inhabitant.
%The first step consists in showing that these types are propositions.
In comparison, our approach
% Actually, we take a different approach for the relation on types
% (and terms) which
yields a more concise definition of the relation.
For example, in the case of the universe, this would lead
to the definition
  $  \Uw\,\wf{\GG} \sim \model{A} := (\wf{\GG}\sim\model{\GG}) \times (\model{A} = \model{\U})$,
  instead of
  our definition
  $  \Uw\,\wf{\GG} \sim \model{A} := (\model{A} = \model{\U})$.
% Recall that our adapted goal, for types, is to prove that
% $\sum_{\model{A}}A\sim\model{A}$ is a proposition, and
% is inhabited if $\GG \sim \model{\GG}$ is.
%  This means that in the definition of the relation, we assume
% that contexts are already related, so that we don't need to enforce them to be,
% as it is the case in the original approach.
% We give the two versions of the universe
% case below, to explain the differences.

@@

@@

  & \rlap{(1) Substitution calculus} \\[0.5em]

@@

  & \rlap{(2) Universe} \\[0.5em]

@@

  & \rlap{(3) Inductive parameters} \\[0.5em]

@@

  & \rlap{(4) Metatheoretic parameters} \\[0.5em]

@@

% As discussed above, when writing the definition of the type components, we assume that the input semantic context is already
% related to the typing judgment of the syntactic context.
% The first version
% of the relation that we suggested at the beginning of the section
% would rather enforce them to be related.
% In the case of $\U$, this would lead
% to the definition
%   $  \Uw\wf{\GG} \sim \model{A} := (\wf{\GG}\sim\model{\GG}) \times (\model{A} = \model{\U})$,
%   instead of
%   the current definition
%   $  \Uw\wf{\GG} \sim \model{A} := (\model{A} = \model{\U})$.
%   Our choice makes the definitions more concise, and similarly
%  in the definition of the term components we assume that the input semantic
%  context and type are already related to their associated well-formedness judgments.
 % However, for some fields, we cannot avoid the first approach.
  % Unfortunately, we need sometimes to require some equalities that are actually
  % redundant with this external assumption.
  %, although we know that these equalities are
  % already satisfied by our assumption.
  % As we argued before, in the  of empty substitution $\epsilon$: we require the equality
  % $(e_C : \model{E} = \model{\boldsymbol{\cdot}})$, although
  % our claimed external assumption that $\model{E}$ is  related to the
  % canonical proof $\epsilonw$ of the typing judgment $\epsilon \vdash$ should imply
  % it.
% Let us comment the first variable case:
  % Another example is the first variable case, where we follow the first verbose
  % approach:
  % the relation gives a semantic context $\model\GG$ that must be related
  % to the syntactic $\GG$, and a semantic
  % type $\model{A}$ related to the syntactic $A$. The input semantic context is
  % then required to equal $\model{\GG}\model{\ext}\model{A}$.   As
  % $\blank \model{\ext}\blank$ is not guaranteed to be injective, these
  % relations are required to show that the relation is right unique.
Next, we prove that this relation is right unique.  Then, we show that the
relation is stable under weakening and substitution.  The last step consists of
showing left-totality, i.e.\ giving a related semantic counterpart to any
well-typed context, type or term.  Everything is proved by induction on the typing
judgments.
  \subsubsection{Right uniqueness}
  \label{sec:right_uniqueness}
  We show by induction that the relation is right unique in the following sense:
    \isp{\Sigma{\sim}}
    && : (\wf\GG : \GG \vdash ) && \ra
    \isaprop\, (\sum_{\model{\GG}} \wf{\GG} \sim \model{\GG})
    \isp{\Sigma{\sim}}
    && : (\wf A : \GG \vdash A) && \ra
    \isaprop\, (\sum_{\model{A}} \wf{A} \sim \model{A})
    \isp{\Sigma{\sim}}
    && : (\wf t : \GG \vdash t \in A) && \ra
    \isaprop\, (\sum_{\model{t}} \wf{t} \sim \model{t})
    \isp{\Sigma{\sim}}
    && : (\wf x : \GG \vdash x \invar A) && \ra
    \isaprop\, (\sum_{\model{x}} \wf{x} \sim \model{x})
    \isp{\Sigma{\sim}}
    && : (\wf \sigma : \GG \vdash \sigma \Rightarrow \GD) && \ra
    \isaprop\, (\sum_{\model{\sigma}} \wf{\sigma} \sim \model{\sigma})
  We mentioned that in order to avoid a recursive-recursive definition, we
  removed some hypotheses in the list of arguments of the relation. Such
  hypotheses are sometimes missed, for example in the case of the empty
  substitution or in the case of variables, requiring us to state additional
  equalities. Because of this, we need UIP to show that
  $\sum_{\model{\GG}}\GG\sim\model{\GG}$ and
  $\sum_{\model{A}}A\sim\model{A}$ are propositions.  One may think that the use
  of UIP could be avoided by using the alternative verbose definition that we
  suggested before, expecting that
  $\sum_{\model{\GG}}\sum_{\model{A}}A\sim\model{A}$, rather than
  $\sum_{\model{A}}A\sim\model{A}$, is a proposition.  However, this is not
  obvious. For example, we were not able to define
  $\Elw\,\wf{\GG}\,\wf{a}\sim \model{A}$ in this fashion. In related work,
  Hugunin investigated constructing IITs without
  UIP \cite{jasper} in cubical type theory, and demonstrated
  that well-formedness predicates used in syntactic models can subtly break
  in that setting. Also, while Hugunin does not use UIP, he only shows a
  weak version of dependent elimination for the constructed IITs. Hence, the
  question whether IITs are reducible to inductive types in a UIP-free setting
  remains open.

@@

  % \todo[inline]{András: Hugunin's paper has an example (which is
  %    likely generalizable) that a notion of redundancy in
  %    well-formedness relations implies that UIP is required.}
  \subsubsection{Stability under weakening}
  % Let us argue that it is necessary to show s
  Stability of the relation under
  weakening must be proved before stability under substitution.
  Indeed, in the proof of stability under substitution, the $\Pi$ case
  requires to show that $\Pi \,(A[\sigma]) \,(B[\keep\,\sigma])$ is related to
  $\model{\Pi}\,(\model{A}\model{[\sigma]})\,
  (\model{B}\model{[\model{\keep}\,\sigma]})$.
  We would like to apply the induction hypothesis, so we need to show that
  $\keep\,\sigma=\varu\,\mathsf{0}\consu\wkO\,\sigma$ is related to
  $\model{\keep}\,\model{\sigma}$, knowing that $\sigma$ is
  related to $\model{\sigma}$.
  As $\keep\,\sigma=\varu\, \mathsf{0}\consu \wkO\,\sigma$, we are left with showing that
  $\wkO\,\sigma=\sigma\circ\wk$ (where $\wk=\wkO\,\id$)
  relates to its semantic counterpart.

@@

\subsubsection{Stability under substitution}

>>>>

We construct the theory of IIT signatures in the following steps:
\begin{enumerate}
\item We view the theory of signatures as a type theory, and we define
  its untyped syntax as mutual inductive types together with typing
  judgments given by inductive relations on the untyped syntax. Then
  the syntax $\I : \SignAlg$ is constructed using those untyped terms
  for which the typing relation holds.
\item We construct $\ll\blank\rr:\SignMor\,\I\,M$ for arbitrary
  $M : \SignAlg$, by:
  \begin{enumerate}
  \item defining a relation $\blank \sim \blank$ between the well-typed syntax
    and a given model. The idea is that given a syntactic context $\GG$ and a
    semantic context $\model{\GG}$ of the model $M$, we have
    $\GG \sim \model{\GG}$ if and only if $\ll\GG\rr
    = \model{\GG}$, and similarly for types, terms, and substitutions;
    % semantic context $\model{\GG}$ of
    % the model $M$ if and only if $\GG$ relates to $\rec\,{\GG}$;
    % enjoyed by the initial morphism from the
    % syntax to a given model;
  \item showing that this relation is functional and thus obtaining a function.
    \end{enumerate}
  \item Proving the uniqueness of this morphism by showing that any morphism
    $f:\SignMor\,\I\,M$ satisfies the relation. For example, for any syntactic context
    $\GG$ we have $\GG\sim f\,\GG$.
      % TODO donner composantes sur les contextes
\end{enumerate}
The next sections detail each of these steps.

@@

Crucially, we do not have conversion relations for typed syntax, nor
do we need to use quotients when giving the syntactic model. This is
possible because there are no $\beta$-rules in the theory of
signatures. Hence, we consider only normal terms in the untyped
syntax, and define weakening and substitution by recursion.  Avoiding
quotients is important for two reasons. First, it greatly simplifies
formalisation. Second, we aim to reduce the theory of signatures only
to inductive types, thus making \Cref{thm:WToToS} stronger.

@@

\subsubsection{Untyped Syntax and its Properties}

@@

\subsubsection{Typing Relations and Their Properties}

\begin{definition}[Typing relations]\label{def:typing_judgments}
  The typing relations are defined as the following inductive type indexed over the
  untyped syntax:
\begin{alignat*}{5}
  & \rlap{(1) Substitution calculus} \\[0.5em]
  & \blank \vdash && : \Conu\ra \Set \\
  & \blank \vdash \blank  && : \Conu\ra \Tyu\ra\Set \\
  & \blank \vdash \blank \invar \blank  && : \Conu\ra \N \ra \Tyu\ra \Set \\
  & \blank \vdash \blank \in \blank  && : \Conu\ra \Tmu\ra \Tyu\ra \Set \\
  & \blank \vdash \blank \Rightarrow \blank  && : \Conu\ra \Subu \ra \Conu\ra \Set \\
  % & \Subpre  && : \Set \\
  & \emptyw && : \emptyu \vdash \\
  & \epsilonw && : \GG\vdash \epsilonu \Rightarrow \emptyu \\
  & \blank\extw\blank && : ( \GG\vdash )  \ra  (\GG\vdash A)  \ra
  \GG\extu A \vdash \\
  & \consw && :
    (\GD \vdash) \ra
    (\GG\vdash \sigma \Rightarrow\GD)\ra
    ( \GD \vdash A) \ra
    ( \GG \vdash t \in A[\sigma]) \ra
    \GG \vdash \sigma\consu t \Rightarrow \GD\extu A
   \\
  & \varw  && : (\GG \vdash n \invar A) \ra \GG \vdash \varu n \in A \\
  & \vzw  && : (\GG\vdash)\ra(\GG\vdash A)\ra\GG \extu  A \vdash \mathsf{0} \invar \wku \, A \\
  & \vsw  && : (\GG\vdash)\ra(\GG\vdash A)\ra(\GG \vdash n \invar A) \ra (\GG \vdash B) \ra \GG \extu  B
  \vdash \mathsf{S}\,n \invar \wku \, A \\
  & \rlap{(2) Universe} \\[0.5em]
  & \Uw && : (\GG \vdash)\ra \GG \vdash \Uu \\
  & \Elw && : (\GG \vdash)\ra (\GG \vdash a \in \Uu)\ra\GG \vdash \Elu\,a \\
  & \rlap{(3) Inductive parameters} \\[0.5em]
\  & \Piw && :
    (\GG \vdash)\ra (\GG \vdash a \in \Uu)\ra(\GG \extu
  \Elu\,a \vdash B)\ra\GG \vdash \Piu \, a \, B \\
  & \appw && :
    (\GG \vdash)\ra (\GG \vdash a \in \Uu)\ra(\GG \extu
    \Elu\,a \vdash B)
    \\
    & &&
    \ra(\GG \vdash t \in \Piu \, a \, B )
    \ra
    (\GG \vdash u \in \Elu\,a)
    \ra
    \GG \vdash t \appu u \in  B [0 := u] \\
  & \rlap{(4) External parameters} \\[0.5em]
  & \Pimw && :
    (T:\Set)\ra(A : T\ra \Tyu)\ra(\GG \vdash)\ra
    ((t : T)\ra\GG\vdash A\,t) \ra
    \GG \vdash \Pimu \,T\,A
    \\
    & \appmw  && :
    (T:\Set)\ra(A : T\ra \Tyu)\ra(\GG \vdash)\ra
    ((t : T)\ra\GG\vdash A\,t)
    \\
    & &&
    \ra(\GG \vdash t \in \Pimu \,T\,A)
    \ra (u : T) \ra \GG \vdash t\,\appmu\,u \in A\,u
  \end{alignat*}
\end{definition}
There is possibility of redundancy in the arguments of the
constructors. Here, we are ``{paranoid}'' (nomenclature from
\cite{paranoid}), so that we get more inductive hypotheses when
performing recursion.

@@

\subsubsection{The Syntax as a Signature Algebra}

@@

\subsection{Relating the Syntax to a Signature Algebra}
\label{sec:streicher}

@@

Given an $M : \SignAlg$, we define the functional relation satisfied by
the  $\ll\blank\rr : \SignMor\,\I\, M$ by recursion on the typing judgments.
% As these judgments are propositions, we will sometimes
If $\GG$ is a context in $\I$ and $\model{\GG}$ is a semantic context
(i.e.\ a context in the signature algebra $M$), we want to define a type
$\GG\sim \model{\GG}$ equivalent to $\ll\GG\rr =\model{\GG}$. Of
course, at this stage, $\ll\blank\rr$ is not available yet since the point of
defining this relation is to construct $\ll\blank\rr$ in the end.

For a type $A$ in a context $\GG$, we want to define a relation
$A\sim \model{A}$ that is equivalent to $\ll A\rr=\model{A}$.
% where $\model{A}$ is a
% semantic type in a semantic context $\model{\GG}$.
For this equality to make sense, the semantic type $\model{A}$ must live
in the semantic context $\ll\GG\rr$. But again, $\ll\blank\rr$ is not
yet available at this stage. Exploiting the expected equivalence between
$\GG\sim\model\GG$ and $\ll\GG\rr=\model\GG$, we may consider defining $A\sim\model{A}$ under the
hypotheses that $\model{A}$ lies in a semantics context $\model\GG$ which is
related to $\GG$. Then, the type of the relation for types is
\[
  (\GG : \syn\Con)\ra (A : \syn\Ty\,\GG) \ra
  (\model\GG : \model\Con)
  \ra
  (\GG \sim \model\GG)
  \ra
  (\model{A}:\model\Ty\,\model\GG)
  \ra
  \Set
\]
Note that the relation on contexts
must be defined mutually with the relation on types (see for example the case of
context extension), but here, the relation on contexts appears as the type of an
argument of the relation on types.
% We leave this possibility aside,
We want to avoid using such recursive-recursive
definitions as they are not allowed by the elimination principles of indexed inductive types, so we instead just remove the hypothesis
$\GG\sim\model\GG$ from the list of arguments.  We proceed similarly for
terms and substitutions. Actually, this removal is not without harm. For
example, consider relating the empty substitution
$\GG\vdash \epsilonu \Rightarrow \emptyu $ to a semantic substitution
$\model\sigma : \model\Sub\,\model\GG\,\model\GD$. We would like to assert
that $\model\sigma$ equals the empty semantic substitution $\model\epsilon$, but
this is not possible because typechecking requires that $\model\GD$ is the
empty semantic context. This is precisely what was ensured by the hypothesis
$\syn\cdot \sim \model\GD$ we removed.  Our way out here is to state that $\model\sigma$
is related to the empty substitution if the target semantic context $\model\GD$
is empty, and, acknowledging this equality, if $\model\sigma$ is the empty
substitution.
% The idea is that we will first prove that there is atmost one semantic type
% which relates to the syntactic one, and then later we will provide such a
% semantic type under the hypothesis that the contexts are related.

Let us mention another possible solution for avoiding recursion-recursion:
defining $A\sim\model{A}$ so that it is equivalent to $(e:\ll\GG\rr
= \model{\GG})\times (\ll A\rr = _{e\#}\model{A})$.
% whereas the right hand side is of type $\model\Ty\model{\GG}$.
% For a substitution $\sigma$ between contexts $\GG$ and $\GD$,
% a first idea consists in defining a relation $\sigma \sim_{\blank} \model{\sigma} : \GG
% \sim \model\GG \to \Set $ where $\model{A}$ is a semantic type in the
% semantic context $\model{\GG}$. We define the relation on types under the hypothesis
% that the contexts are already related.
% it is natural to define the relation with a semantic type $\model{A}$ in a
% semantic context $\model{\GG}$ such that it is equivalent to  $(\rec\,\GG = \model{\GG})\times (\rec\,A = \model{A})$ (note that the first equality is
% required for the second one to be well-typed).
% The discussion is similar for terms.
% Then, the ultimate goal is to prove that
% $\sum_{\model{\GG}}\GG\sim\model{\GG}$ and
% $\sum_{\model{\GG}}({\model{A}:\model{\Ty}\,\model{\GG}})\times(A\sim\model{A})$
% are contractible, i.e., have a unique inhabitant.
%The first step consists in showing that these types are propositions.
In comparison, our approach
% Actually, we take a different approach for the relation on types
% (and terms) which
yields a more concise definition of the relation.
For example, in the case of the universe, this would lead
to the definition
  $  \Uw\,\wf{\GG} \sim \model{A} := (\wf{\GG}\sim\model{\GG}) \times (\model{A} = \model{\U})$,
  instead of
  our definition
  $  \Uw\,\wf{\GG} \sim \model{A} := (\model{A} = \model{\U})$.
% Recall that our adapted goal, for types, is to prove that
% $\sum_{\model{A}}A\sim\model{A}$ is a proposition, and
% is inhabited if $\GG \sim \model{\GG}$ is.
%  This means that in the definition of the relation, we assume
% that contexts are already related, so that we don't need to enforce them to be,
% as it is the case in the original approach.
% We give the two versions of the universe
% case below, to explain the differences.

@@

\begin{definition}[Relation $\blank\sim\blank$]

@@

  & \rlap{(1) Substitution calculus} \\

@@

  & \rlap{(2) Universe} \\

@@

  & \rlap{(3) Inductive parameters} \\

@@

  & \rlap{(4) Metatheoretic parameters} \\

@@

\end{definition}

% As discussed above, when writing the definition of the type components, we assume that the input semantic context is already
% related to the typing judgment of the syntactic context.
% The first version
% of the relation that we suggested at the beginning of the section
% would rather enforce them to be related.
% In the case of $\U$, this would lead
% to the definition
%   $  \Uw\wf{\GG} \sim \model{A} := (\wf{\GG}\sim\model{\GG}) \times (\model{A} = \model{\U})$,
%   instead of
%   the current definition
%   $  \Uw\wf{\GG} \sim \model{A} := (\model{A} = \model{\U})$.
%   Our choice makes the definitions more concise, and similarly
%  in the definition of the term components we assume that the input semantic
%  context and type are already related to their associated well-formedness judgments.
 % However, for some fields, we cannot avoid the first approach.
  % Unfortunately, we need sometimes to require some equalities that are actually
  % redundant with this external assumption.
  %, although we know that these equalities are
  % already satisfied by our assumption.
  % As we argued before, in the  of empty substitution $\epsilon$: we require the equality
  % $(e_C : \model{E} = \model{\boldsymbol{\cdot}})$, although
  % our claimed external assumption that $\model{E}$ is  related to the
  % canonical proof $\epsilonw$ of the typing judgment $\epsilon \vdash$ should imply
  % it.
% Let us comment the first variable case:
  % Another example is the first variable case, where we follow the first verbose
  % approach:
  % the relation gives a semantic context $\model\GG$ that must be related
  % to the syntactic $\GG$, and a semantic
  % type $\model{A}$ related to the syntactic $A$. The input semantic context is
  % then required to equal $\model{\GG}\model{\ext}\model{A}$.   As
  % $\blank \model{\ext}\blank$ is not guaranteed to be injective, these
  % relations are required to show that the relation is right unique.

\subsubsection{Right Uniqueness}

Next, we prove that this relation is right unique.  Then, we show that the
relation is stable under weakening and substitution.  The last step consists of
showing left-totality, i.e.\ giving a related semantic counterpart to any
well-typed context, type or term.  Everything is proved by induction on the typing
judgments.

\begin{lemma}[Right uniqueness]\label{lem:right_uniqueness}
The relation is right unique in the following sense:
  \begin{alignat*}{5}
    &
    \isp{\Sigma{\sim}}
    && : (\wf\GG : \GG \vdash ) && \ra
    \isaprop\, (\sum_{\model{\GG}} \wf{\GG} \sim \model{\GG})
    \\
    &
    \isp{\Sigma{\sim}}
    && : (\wf A : \GG \vdash A) && \ra
    \isaprop\, (\sum_{\model{A}} \wf{A} \sim \model{A})
    \\
    &
    \isp{\Sigma{\sim}}
    && : (\wf t : \GG \vdash t \in A) && \ra
    \isaprop\, (\sum_{\model{t}} \wf{t} \sim \model{t})
    \\
    &
    \isp{\Sigma{\sim}}
    && : (\wf x : \GG \vdash x \invar A) && \ra
    \isaprop\, (\sum_{\model{x}} \wf{x} \sim \model{x})
    \\
    &
    \isp{\Sigma{\sim}}
    && : (\wf \sigma : \GG \vdash \sigma \Rightarrow \GD) && \ra
    \isaprop\, (\sum_{\model{\sigma}} \wf{\sigma} \sim \model{\sigma})
  \end{alignat*}
\end{lemma}

\begin{remark}
  We mentioned that in order to avoid a recursive-recursive definition, we
  removed some hypotheses in the list of arguments of the relation. Such
  hypotheses are sometimes missed, for example in the case of the empty
  substitution or in the case of variables, requiring us to state additional
  equalities. Because of this, we need UIP to show that
  $\sum_{\model{\GG}}\GG\sim\model{\GG}$ and
  $\sum_{\model{A}}A\sim\model{A}$ are propositions.  One may think that the use
  of UIP could be avoided by using the alternative verbose definition that we
  suggested before, expecting that
  $\sum_{\model{\GG}}\sum_{\model{A}}A\sim\model{A}$, rather than
  $\sum_{\model{A}}A\sim\model{A}$, is a proposition.  However, this is not
  obvious. For example, we were not able to define
  $\Elw\,\wf{\GG}\,\wf{a}\sim \model{A}$ in this fashion. In related work,
  Hugunin investigated constructing IITs without
  UIP \cite{jasper} in cubical type theory, and demonstrated
  that well-formedness predicates used in syntactic models can subtly break
  in that setting. Also, while Hugunin does not use UIP, he only shows the simple version
  version of dependent elimination for the constructed IITs. Hence, the
  question whether IITs are reducible to inductive types in a UIP-free setting
  remains open.
\end{remark}

\subsubsection{Stability under Weakening and Substitution}

@@

  % \todo[inline]{András: Hugunin's paper has an example (which is
  %    likely generalizable) that a notion of redundancy in
  %    well-formedness relations implies that UIP is required.}
  % Let us argue that it is necessary to show s
  Stability of the relation under
  weakening must be proved before stability under substitution.
  Indeed, in the proof of stability under substitution, the $\Pi$ case
  requires to show that $\Pi \,(A[\sigma]) \,(B[\keep\,\sigma])$ is related to
  $\model{\Pi}\,(\model{A}\model{[\sigma]})\,
  (\model{B}\model{[\model{\keep}\,\sigma]})$.
  We would like to apply the induction hypothesis, so we need to show that
  $\keep\,\sigma=\wkO\,\sigma\consu \varu\,\mathsf{0}$ is related to
  $\model{\keep}\,\model{\sigma}$, knowing that $\sigma$ is
  related to $\model{\sigma}$.
  As $\keep\,\sigma=\wkO\,\sigma\consu \varu\, \mathsf{0}$, we are left with showing that
  $\wkO\,\sigma=\sigma\circ\wk$ (where $\wk=\wkO\,\id$)
  relates to its semantic counterpart.

@@

Next, we want to prove that given any well-typed substitution $\sigma
: \Sub\,\GG\,\GD$ and semantic contexts $\model{\GG}$ and
$\model{\GD}$, related to $\GG$ and $\GD$, respectively, there is a
semantic substitution related to $\sigma$.  In the extension case
$\GG\vdash \sigma \consu\,t\Rightarrow \GD\,\extu\,A$, the induction
hypothesis provides $\model{\sigma}$, $\model{\GD}$, $\model{A}$ related to
their syntactic counterpart. However, the premises of the induction hypothesis
for getting a relevant $\model{t}$ require showing that the type
$\model{A}\model{[\model{\sigma}]}$ is related to the syntactic type
$A[\sigma]$.
\begin{lemma}[Preservation of the relation by substitution for variables]
\begin{alignat*}{5}
    % []V~
  & []{\sim} && :
  (\wf{\sigma} : \GG \vdash \sigma \Rightarrow \GD) \ra
  (\wf\sigma \sim \model{\sigma}) \ra
  (\wf{x} : \GD \vdash x \invar A) \ra
  (\wf{x} \sim \model{x}) \ra
  \\ & && \qquad
  % \\
  % & && \qquad
   \wf{[]}\wf{x}\wf{\sigma} \sim \model{x}\model{[\model{\sigma}]}
\end{alignat*}
\end{lemma}
\begin{proof}
  Induction on typing.
\end{proof}
